---
title: "Course Project Statistical Inference, Part 1"
author: 'Author: Hendrik L.'
date: "August 2015"
output: pdf_document
---

```{r setoptions, echo=FALSE, results='hide', message=FALSE}
# knitr library for processing the present R markdown
library(knitr)
opts_chunk$set(echo = FALSE, results = "hide", cache = FALSE)

# ggplot2 library for enhanced plotting
library(ggplot2)
```

### Overview

This project part investigates the exponential distribution in R and compares it with the Central Limit Theorem. The exponential distribution can be simulated with rexp(n, lambda) where lambda is set to `lambda = 0.2` for all simulations. Focus is to investigate the distribution of averages of 40 exponentials by means of 1,000 simulations. 

### Simulation

Exponential distributions can be simulated in R with `rexp(n, lambda)` where `lambda` is the rate parameter. Both the mean and the standard deviation of the exponential distribution are: `1/lambda`

For all simulations, lambda is set to `lambda = 0.2`. The following 1,000 simulations in R are run in order to investigate the distribution of averages of 40 exponentials:

```{r, echo = TRUE, results = 'markup'}
set.seed(164720)
lambda = 0.2
exponentials = 40
simulationMeans = vector()
for (i in 1:1000) { simulationMeans = c(simulationMeans, mean(rexp(exponentials, lambda))) }
```

### 1. Sample mean versus theoretical mean

### 1.1. Sample mean
Calculation of the mean from the simulations which is the sample mean:
```{r, echo = TRUE, results = 'markup'}
mean(simulationMeans)
```

### 1.2. Theoretical mean
The theoretical mean of an exponential distribution is: `lambda^-1`
```{r, echo = TRUE, results = 'markup'}
lambda^-1
```

### 1.3. Conclusion
There is only a small difference between the mean of the simulation samples and the theoretical mean of the exponential distribution.
```{r, echo = TRUE, results = 'markup'}
abs(mean(simulationMeans) - lambda^-1)
```

### 2. Sample variance versus theoretical Variance
### 2.1. Sample variance
```{r, echo = TRUE, results = 'markup'}
var(simulationMeans)
```

### 2.2. Theoretical variance
The theoretical variance of the exponential distribution is: `(lambda * sqrt(n))^-2`
```{r, echo = TRUE, results = 'markup'}
(lambda * sqrt(exponentials))^-2
```

### 2.3. Conclusion
There is only a small difference between the variance of the simulation samples and the theoretical variance of the exponential distribution.
```{r, echo = TRUE, results = 'markup'}
abs(var(simulationMeans) - (lambda * sqrt(exponentials))^-2)
```

### 3. Distribution
The following plot shows a histogram of the 1,000 simulations. In an overlay, also the normal distribution is shown that has a mean of `lambda^-1` and a standard deviation of `(lambda * sqrt(n))^-1` which is the theoretical normal distribution for the simulations. As expected according to the Central Limit Theorem, the 1,000 simulations of  averages of 40 exponentials are approximately normally distributed.
```{r, echo = TRUE, results = 'markup'}
ggplot(data.frame(y = simulationMeans), aes(x = y)) + 
    geom_histogram(aes(y = ..density..), binwidth = 0.2, 
        fill = "green", color = "black") +
    stat_function(fun = dnorm, arg = list(mean = lambda^-1, 
    sd = (lambda * sqrt(exponentials))^-1), size = 2) +
    labs(title = "Simulation means vs. normal distribution", x = "Simulation means")
```
  
&nbsp;  
  
&nbsp;  
  